An Empirical Study of Invasion and Resistance for Iterated Prisoner's Dilemma Strategies

### 2017-05-13 main.tex
Add fsm imgs, algs and fix naming confusion.
I had mixed up TF1 and TF3. This is now fixed throughout.

% Packages
\usepackage[margin=1.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{hyperref}
\usepackage{multicol}
\usepackage{booktabs}
\usepackage{xstring}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{standalone}
\usepackage{fancyvrb}

@@

There are three further strategies trained specifically for this study; Trained
FSM 1, 2, and 3 (TF1 - TF3). These are based on finite state machines of 8, 16,
and 16 states respectively (see Figures \ref{}).
% TODO Include images of FSMs
As opposed to the previously described strategies, these strategies were trained
with the objective function of \textbf{mean fixation probabilities for Moran
processes} starting at initial population states consisting of \(N/2\)
individuals of the training candidates and \(N/2\) individuals of an opponent
strategy, taken from a selection of 150 opponents from the axelrod library:

\begin{itemize}
	\item TF1 \(N=8\), 1\% noise, 100 repetitions per match
	\item TF2 \(N=10\), 0\% noise, 10000 repetitions per match
	\item TF3 \(N=12\), 0\% noise, 10000 repetitions per match
\end{itemize}

@@

TF1 cooperates and defects with
various cycles depending on the opponent's actions. TF1 will mutually
cooperate with any strategy and only tolerates a few defections before
defecting for the rest of match. It is similar to but not exactly the same as
Fool Me Once, a strategy that cooperates until the opponent has defected twice
(not necessarily consecutively), and defects indefinitely thereafter. Though a
product of training with a Moran objective. Interestingly it has no handshake
mechanism, this could be due to the fact that it is trained in a noisy
environment.

@@

TF3 has an initial handshake of CCD and cooperates if the opponent matches.
However if the opponent later defects, TF3 will respond in kind, so the
handshake is not permanent. Only one player (Prober 4 \cite{prison}) manages to
achieve cooperation with TF3 after about 20 rounds of play.

For both TF2 and TF3 a handshake
mechanism naturally emerges from the structure of the underlying finite state
machine. This behavior is an outcome of the evolutionary process and is in no
way hard-coded or included via an additional mechanism.

@@

Each strategy pair is run for 1000 repetitions of the Moran process to fixation
with starting population distributions of $(1, N-1)$, $(N/2, N/2)$ and $(N-1 ,
1)$, for \(N\) from 2 through 14. The fixation probability is then empirically
computed for each combination of starting distribution and value of \(N\).  The
axelrod library can carry out exact simulations of the Moran process. Since some
of the strategies have a high computational cost or are stochastic, samples are
taken
from a large number of match outcomes for the pairs of players for use in
computing fitnesses in the Moran process. This approach was verified to agree
with unsampled calculations to a high degree of accuracy in specific cases.
This is described in Figure~\ref{alg:psuedocode}.
% TODO Include pseudo code

@@

\begin{itemize}
    \item Grudger (which only performs well for \(N=3\)), starts by cooperating
        but will defect if at any point the opponent has defected.
    \item MEM2, an infinite memory strategy that switches between TfT, Tf2T, and
        Defector \cite{Li2014}.
    \item TF1, the finite state machine trained specifically for Moran processes
        described in Section~\ref{sec:introduction}.
    \item Prober 4, complex strategy with an initial 20 move sequence of
        cooperations and defections \cite{prison}. This initial sequence serves
        as approximate handshake.
    \item  PSO Gambler and Evolved Lookerup 2 2 2: are strategies that make use
        of a lookup table mapping the first 2 moves of the opponent as well as
        the last 2 moves of both players to an action. The PSO gambler is a
        stochastic version which maps those states to probabilities of
        cooperating. The lookerup was described in \cite{Knight2016}.
	\item The evolved ANN strategies are neural networks that map a number of
		attributes (first move, number of cooperations, last move etc\dots) to
		an action. Both of these have been trained using an evolutionary
		algorithm and the ANN 5 was trained to perform well in a noisy
		tournament.
    \item The Evolved FSM 16 is a 16 state finite state machine trained to
        perform well in tournaments.
\end{itemize}

As well as noting that the memory length and complexity of these strategies are
much greater than one, it is interesting to note that none of them are akin to
memory one strategies. Only one is stochastic although close inspection of the
source code of PSO Gambler shows that it makes stochastic decisions with extreme
probabilities: almost always acting as expected. Apart from TF1 in \(N=3\), the
finite state machines trained specifically for Moran processes do not appear in
the top 5: whereas strategies trained for tournament do. This is due to the
nature of invasion: most of the opponents will initially be different
strategies. The next section will consider the converse situation.

@@

There are are only two new strategies that appear in the top ranks for
\(x_{N-1}\): TF2 and TF3. These two strategies are with CS the strongest
resistors. They all have handshakes, and whilst the handshake that CS and
Handshake (which ranks highly for the smaller values of \(N\)) have was
programmed, the handshakes of TF2 and TF3 evolved through an evolutionary
process.

@@

A handshake requires at least one defection and there is
selective pressure to defect as few times as possible to achieve the
self-recognition mechanism. It is also unwise to defect on the first move as
some strategies additionally retaliate first round defections. So the
handshakes used by TF2 and TF3, and CS, are in some sense optimal. These
discoveries may have significant ramifications regarding the evolution of
cooperation and forgiveness in biological organisms such as antibacterial
resistant bacteria and social interactions
between humans.

@@

One of the major findings discussed in Section~\ref{sec:strong_resistors}, is
the ability of strategies with a handshake mechanism to resist invasion. This
was not only identify for CS (an intelligently designed strategy) but also for
two FSM strategies (TF2 and TF3) specifically trained through an evolutionary
process. In these two cases, the handshake mechanism was a product of the
evolutionary process. This has the potential to help with the understanding of
organisms with a strong resistance to invasion such as anti antibacterial
resistant bacteria \cite{Davies2010}. With the knowledge that a handshake being
likely to exist, perhaps it can be mimicked.

>>>>

% Packages
\usepackage[margin=1.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{hyperref}
\usepackage{multicol}
\usepackage{booktabs}
\usepackage{xstring}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{standalone}
\usepackage{fancyvrb}
\usepackage{algorithm,algorithmic}

@@

There are three further strategies trained specifically for this study; Trained
FSM 1, 2, and 3 (TF1 - TF3). These are based on finite state machines of 16, 16,
and 8 states respectively (see Figures~\ref{fig:tf1},~\ref{fig:tf2}
and~\ref{fig:tf3}).

\begin{figure}[!hbtp]
    \centering
    \input{./tex/fsm_one.tex}
    \caption{TF1: a 16 state finite state machine.}
    \label{fig:tf1}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \input{./tex/fsm_two.tex}
    \caption{TF2: a 16 state finite state machine.}
    \label{fig:tf2}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \input{./tex/fsm_three.tex}
    \caption{TF3: an 8 state finite state machine.}
    \label{fig:tf3}
\end{figure}

As opposed to the previously described strategies, these strategies were trained
with the objective function of \textbf{mean fixation probabilities for Moran
processes} starting at initial population states consisting of \(N/2\)
individuals of the training candidates and \(N/2\) individuals of an opponent
strategy, taken from a selection of 150 opponents from the axelrod library:

\begin{itemize}
	\item TF1 \(N=12\), 0\% noise, 10000 repetitions per match
	\item TF2 \(N=10\), 0\% noise, 10000 repetitions per match
	\item TF3 \(N=8\), 1\% noise, 100 repetitions per match
\end{itemize}

@@

TF3 cooperates and defects with
various cycles depending on the opponent's actions. TF3 will mutually
cooperate with any strategy and only tolerates a few defections before
defecting for the rest of match. It is similar to but not exactly the same as
Fool Me Once, a strategy that cooperates until the opponent has defected twice
(not necessarily consecutively), and defects indefinitely thereafter. Though a
product of training with a Moran objective. Interestingly it has no handshake
mechanism, this could be due to the fact that it is trained in a noisy
environment.

@@

TF1 has an initial handshake of CCD and cooperates if the opponent matches.
However if the opponent later defects, TF1 will respond in kind, so the
handshake is not permanent. Only one player (Prober 4 \cite{prison}) manages to
achieve cooperation with TF1 after about 20 rounds of play.

For both TF1 and TF2 a handshake
mechanism naturally emerges from the structure of the underlying finite state
machine. This behavior is an outcome of the evolutionary process and is in no
way hard-coded or included via an additional mechanism.

@@

Each strategy pair is run for 1000 repetitions of the Moran process to fixation
with starting population distributions of $(1, N-1)$, $(N/2, N/2)$ and $(N-1 ,
1)$, for \(N\) from 2 through 14. The fixation probability is then empirically
computed for each combination of starting distribution and value of \(N\).  The
axelrod library can carry out exact simulations of the Moran process. Since some
of the strategies have a high computational cost or are stochastic, samples are
taken
from a large number of match outcomes for the pairs of players for use in
computing fitnesses in the Moran process. This approach was verified to agree
with unsampled calculations to a high degree of accuracy in specific cases.
This is described in Algorithms~\ref{alg:data_collection}
and~\ref{alg:moran_process}..

\begin{algorithm}[!hbtp]
        \caption{Data Collection}
        \label{alg:data_collection}
        \input{./tex/data_collection.tex}
\end{algorithm}

\begin{algorithm}[!hbtp]
        \caption{Moran process}
        \label{alg:moran_process}
        \input{./tex/moran_process_simulation.tex}
\end{algorithm}

@@

\begin{itemize}
    \item Grudger (which only performs well for \(N=3\)), starts by cooperating
        but will defect if at any point the opponent has defected.
    \item MEM2, an infinite memory strategy that switches between TfT, Tf2T, and
        Defector \cite{Li2014}.
    \item TF3, the finite state machine trained specifically for Moran processes
        described in Section~\ref{sec:introduction}.
    \item Prober 4, complex strategy with an initial 20 move sequence of
        cooperations and defections \cite{prison}. This initial sequence serves
        as approximate handshake.
    \item  PSO Gambler and Evolved Lookerup 2 2 2: are strategies that make use
        of a lookup table mapping the first 2 moves of the opponent as well as
        the last 2 moves of both players to an action. The PSO gambler is a
        stochastic version which maps those states to probabilities of
        cooperating. The lookerup was described in \cite{Knight2016}.
	\item The evolved ANN strategies are neural networks that map a number of
		attributes (first move, number of cooperations, last move etc\dots) to
		an action. Both of these have been trained using an evolutionary
		algorithm and the ANN 5 was trained to perform well in a noisy
		tournament.
    \item The Evolved FSM 16 is a 16 state finite state machine trained to
        perform well in tournaments.
\end{itemize}

As well as noting that the memory length and complexity of these strategies are
much greater than one, it is interesting to note that none of them are akin to
memory one strategies. Only one is stochastic although close inspection of the
source code of PSO Gambler shows that it makes stochastic decisions with extreme
probabilities: almost always acting as expected. Apart from TF3 in \(N=3\), the
finite state machines trained specifically for Moran processes do not appear in
the top 5: whereas strategies trained for tournament do. This is due to the
nature of invasion: most of the opponents will initially be different
strategies. The next section will consider the converse situation.

@@

There are are only two new strategies that appear in the top ranks for
\(x_{N-1}\): TF1 and TF2. These two strategies are with CS the strongest
resistors. They all have handshakes, and whilst the handshake that CS and
Handshake (which ranks highly for the smaller values of \(N\)) have was
programmed, the handshakes of TF1 and TF2 evolved through an evolutionary
process.

@@

A handshake requires at least one defection and there is
selective pressure to defect as few times as possible to achieve the
self-recognition mechanism. It is also unwise to defect on the first move as
some strategies additionally retaliate first round defections. So the
handshakes used by TF1 and TF2, and CS, are in some sense optimal. These
discoveries may have significant ramifications regarding the evolution of
cooperation and forgiveness in biological organisms such as antibacterial
resistant bacteria and social interactions
between humans.

@@

One of the major findings discussed in Section~\ref{sec:strong_resistors}, is
the ability of strategies with a handshake mechanism to resist invasion. This
was not only identify for CS (an intelligently designed strategy) but also for
two FSM strategies (TF1 and TF2) specifically trained through an evolutionary
process. In these two cases, the handshake mechanism was a product of the
evolutionary process. This has the potential to help with the understanding of
organisms with a strong resistance to invasion such as anti antibacterial
resistant bacteria \cite{Davies2010}. With the knowledge that a handshake being
likely to exist, perhaps it can be mimicked.
